---
title: "API lifecycle and reproducibility contracts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{API lifecycle and reproducibility contracts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why the package has contracts

`SOMevidence` treats software interfaces as part of reproducibility. A change to
an argument name or a returned evidence table can invalidate an analysis even
when the numerical algorithm is unchanged. The package therefore ships two
machine-readable versioned contracts:

```{r contracts}
api_path <- system.file(
  "extdata", "api-contract-v1.0.0.csv", package = "SOMevidence"
)
object_path <- system.file(
  "extdata", "object-contract-v1.0.0.csv", package = "SOMevidence"
)
api_contract <- utils::read.csv(api_path, check.names = FALSE)
object_contract <- utils::read.csv(object_path, check.names = FALSE)

table(api_contract$lifecycle)
head(object_contract[c("class", "lifecycle")])
```

Automated tests compare these files with the namespace, function formals and
returned object structures. The 1.0.0 contracts define the first public API
compatibility promise.

## Stable and experimental interfaces

Stable functions cover the data contract, preprocessing, sampling design,
ensemble fitting, evidence audits, transfer and conditional decision gates.
Breaking changes require a new major package version.

Three interfaces remain experimental:

```{r experimental}
api_contract[api_contract$lifecycle == "experimental",
             c("function", "notes")]
```

The optional GUI exports ordinary R code and a YAML configuration. The R code,
not the interactive session, is the reproducible record. The sensitivity and
simulation interfaces remain experimental because external users may expose
better ways to describe scenarios.

## Object boundaries are scientific boundaries

The returned classes keep distinct questions apart:

- `som_audit` concerns representation quality;
- `som_partitions` and `som_consensus` concern analysis-scope hard-partition
  stability; held-out mappings belong to transfer evidence;
- `som_cross_comparison` concerns algorithmic agreement;
- `som_transfer_audit` concerns held-domain behaviour; and
- `som_defensibility` records an analyst-specified conditional decision.

No object is a certificate of ecological truth. External labels are held out
from training, cross-model agreement is not a leaderboard and insufficient
evidence may lead to abstention.

## Reporting an analysis

Archive the package version, session information, analysis script, sampling
design and gate settings with every analysis:

```{r report-version}
packageVersion("SOMevidence")
sessionInfo()
```

For work intended to support a publication, also retain the exact data
provenance, random seeds, model failures, warnings and the API contract
version used by the analysis.
