---
title: "Respecting groups and monitoring domains in SOM validation"
author: "Shaowen Ye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Respecting groups and monitoring domains in SOM validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why the resampling unit matters

Environmental and ecological tables often contain repeated observations from
the same site, individual, transect, campaign or watershed. Treating those
rows as independent can place closely related observations in both the
analysis and assessment subsets. A model may then appear reproducible because
it repeatedly sees the same sampling unit, not because the pattern transfers
to a new unit.

`SOMevidence` stores the sampling hierarchy outside the predictor matrix and
requires the resampling design to be stated. The example below creates a known
three-class system with strong site-level dependence. Class labels remain
external to training.

```{r grouped-data}
library(SOMevidence)

grouped <- simulate_som_scenario(
  "grouped_pseudoreplication",
  n = 120,
  p = 6,
  n_groups = 20,
  group_icc = 0.7,
  seed = 101
)
grouped
```

Row resampling and site resampling answer different questions. The latter asks
whether a partition can be reproduced after entire sites are withheld.

```{r resampling-designs}
row_splits <- som_resamples(
  grouped, method = "subsample", repeats = 3, prop = 0.8, seed = 102
)
site_splits <- som_resamples(
  grouped, method = "group_subsample", repeats = 3, prop = 0.8, seed = 102
)

spec <- som_spec(
  grids = list(c(3, 2), c(4, 3)),
  seeds = c(103, 104),
  rlen = 40,
  k = 2:4
)
```

Both workflows use the same SOM budget and preprocessing. Only the resampling
unit changes.

```{r grouped-workflows}
row_result <- run_som_workflow(
  grouped, spec, row_splits,
  cross_models = c("kmeans", "ward")
)
site_result <- run_som_workflow(
  grouped, spec, site_splits,
  cross_models = c("kmeans", "ward")
)

row_result$partitions$stability
site_result$partitions$stability
```

These outputs should be compared as design sensitivities, not as an algorithm
ranking. A difference shows that the inferential unit matters. It does not by
itself identify which design matches a particular field study; that decision
comes from how the observations were collected.

## Mapping to a held-out monitoring domain

Leave-domain-out analysis addresses another question: does a map trained in
some domains receive observations from a held-out domain at comparable
distances and on units occupied during training?

```{r transfer-data}
shifted <- simulate_som_scenario(
  "gradient",
  n = 120,
  p = 6,
  n_domains = 3,
  domain_shift = 1,
  seed = 110
)
domain_splits <- som_resamples(
  shifted,
  method = "leave_domain_out",
  domain = "domain"
)
domain_ensemble <- fit_som_ensemble(
  shifted,
  som_spec(c(4, 3), seeds = c(111, 112), rlen = 40, k = 2:4),
  domain_splits
)
transfer <- audit_transfer(domain_ensemble)
transfer
```

```{r transfer-plot, fig.width = 6, fig.height = 4}
plot(transfer)
```

The distance ratio and unoccupied-unit rate diagnose domain shift. They are
not accuracy statistics, because an unsupervised map has no response label to
predict. They also do not establish a causal explanation for the shift. In
this teaching generator, the domains occupy successive ranges of the latent
gradient and `domain_shift` adds a domain-specific offset. The example therefore
combines range extrapolation with an additive shift; it is not an isolated test
of either mechanism.

## Map a later monitoring batch without refitting

An operational monitoring workflow may freeze a trained ensemble and map a
later batch. The new data must have the same named layers and variables, but it
may contain one or many observations. Each map applies its own training-derived
preprocessing.

```{r new-batch}
training_rows <- shifted$metadata$domain != "domain_03"
training <- som_data(
  layers = lapply(
    shifted$layers,
    function(x) x[training_rows, , drop = FALSE]
  ),
  id = shifted$metadata$id[training_rows]
)
new_batch <- som_data(
  layers = lapply(
    shifted$layers,
    function(x) x[!training_rows, , drop = FALSE]
  ),
  id = shifted$metadata$id[!training_rows],
  domain = shifted$metadata$domain[!training_rows]
)
frozen_ensemble <- fit_som_ensemble(
  training,
  som_spec(c(4, 3), seeds = c(120, 121), rlen = 40, k = 2:4),
  keep_models = TRUE
)
new_mapping <- map_som_ensemble(frozen_ensemble, new_batch)
new_mapping
new_mapping$summary
```

```{r new-batch-plot, fig.width = 6, fig.height = 4}
plot(new_mapping)
```

Best-matching unit identifiers stay specific to each map. The package does not
align nodes across different grids or turn distance shift into a hard class
prediction.

## Predefined temporal or spatial blocks

`block_subsample` expects `unit` to identify blocks already defined from the
study design. The package does not guess a block length from row order. This
keeps the scientific definition of a season, campaign, reach or time window
under analyst control and prevents accidental dependence on file ordering.
