---
title: "A real-data audit of Palmer penguin morphology"
author: "Shaowen Ye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A real-data audit of Palmer penguin morphology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_tutorial_dependencies <-
  requireNamespace("palmerpenguins", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_tutorial_dependencies
)
```

## Purpose and inferential boundary

This tutorial uses an openly licensed, familiar dataset to show the complete
workflow on real observations. It does not claim that penguin species are
latent SOM classes or that morphology alone defines species. The four
morphological variables are used for unsupervised training. Species is retained
only as an external label and is inspected after the ensemble and consensus
partition have been constructed.

The data are distributed by the
[`palmerpenguins`](https://allisonhorst.github.io/palmerpenguins/) package under
CC0. Measurements were collected for Adelie, Chinstrap, and Gentoo penguins in
the Palmer Archipelago from 2007 to 2009. The original data publication and
source citations are provided by the data package.

If the optional tutorial dependencies are unavailable, the code is displayed
but not evaluated when the vignette is built.

## Build a design-explicit data object

```{r prepare-data}
library(SOMevidence)

raw <- palmerpenguins::penguins
variables <- c(
  "bill_length_mm", "bill_depth_mm",
  "flipper_length_mm", "body_mass_g"
)
keep <- stats::complete.cases(raw[, c(variables, "species", "year")])
penguins <- raw[keep, , drop = FALSE]

morphology <- som_data(
  x = penguins[, variables],
  id = sprintf("penguin_%03d", seq_len(nrow(penguins))),
  time = penguins$year,
  domain = factor(penguins$year),
  external_label = penguins$species
)
morphology
```

Stable row identifiers make later joins auditable. Sampling year is declared as
a transfer domain. Species is stored in metadata and never enters the SOM input
matrix.

## Prespecify the transfer design and ensemble budget

Each split below leaves out one sampling year. The map and every preprocessing
quantity are fitted using the other years; observations from the held-out year
are assessment data. This demonstration uses a small budget so that the
vignette remains quick to build. A substantive analysis should justify its
grids, seeds, iterations, candidate `k`, and transfer domains before fitting.

```{r design}
year_splits <- som_resamples(
  morphology,
  method = "leave_domain_out",
  domain = "domain"
)

spec <- som_spec(
  grids = list(c(4, 3), c(5, 4)),
  seeds = c(41, 42),
  rlen = 50,
  k = 2:4
)

year_splits
expand_som_spec(spec)
```

## Fit once and retain separate evidence streams

```{r workflow}
workflow <- run_som_workflow(
  morphology,
  spec,
  year_splits,
  preprocess = som_preprocess(center = TRUE, scale = TRUE),
  cross_models = c("kmeans", "ward")
)
workflow
```

The representation audit, partition audit, consensus summaries, cross-model
comparison, and transfer audit answer different questions. They should not be
collapsed into a single rank.

```{r summaries}
workflow$audit$grid_summary
workflow$partitions$stability
workflow$consensus$k3$cluster_summary
workflow$cross_comparison$summary

transfer <- audit_transfer(workflow$ensemble)
transfer$metrics
```

```{r evidence-plots, fig.width = 7, fig.height = 4, out.width = "90%"}
plot(workflow$audit)
plot(workflow$partitions)
plot(transfer)
```

Quantization and topographic errors concern the SOM representation. ARI and AMI
between SOM and reference partitions concern agreement under a shared analysis
design. Held-year distance and occupancy diagnostics concern transfer. None of
these quantities is classification accuracy.

## Inspect external labels only after consensus

For illustration, inspect `k = 3` because the dataset contains three recorded
species. This is a prespecified teaching choice, not evidence that `k = 3` is
the optimal unsupervised partition.

```{r external-labels}
consensus_k3 <- workflow$consensus$k3
external <- evaluate_external_labels(consensus_k3)
external
external$contingency
external$composition
```

The resulting ARI, AMI, and contingency table describe agreement between one
consensus partition and the recorded species labels among evaluable samples.
They do not establish that the partition is biologically correct, and they do
not validate the package. The example instead demonstrates how labels can be
kept outside training and introduced transparently at the interpretation stage.

## Reproducibility record

Archive the package version, software environment, input provenance, and exact
analysis choices with any reported result.

```{r reproducibility}
packageVersion("SOMevidence")
packageVersion("palmerpenguins")
sessionInfo()
```
