---
title: "Start here: a complete SOM evidence workflow"
author: "Shaowen Ye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Start here: a complete SOM evidence workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What this example does

This is the shortest complete route through `SOMevidence`. We will:

1. prepare data and a resampling design;
2. fit a small SOM ensemble;
3. inspect the map and candidate partitions; and
4. compare the SOM partitions with two reference methods.

The example uses simulated classes, so we know how the data were generated.
With field data, the underlying groups are unknown. A stable partition is
evidence worth reporting, but it is not proof of an ecological type.

```{r data}
library(SOMevidence)

dat <- simulate_som_scenario(
  "clusters",
  n = 90,
  p = 6,
  seed = 10,
  id = paste0("example_", seq_len(90))
)
dat
```

The simulated class label is stored in metadata. It does not enter the SOM
training matrix.

## Describe the analysis before fitting

Rows are independent in this simulation, so ordinary subsampling is suitable.
For repeated observations from sites, individuals or transects, resample whole
groups instead. See the resampling tutorial for worked examples.

```{r design}
resamples <- som_resamples(
  dat,
  method = "subsample",
  repeats = 3,
  prop = 0.8,
  seed = 20
)

spec <- som_spec(
  grids = list(c(3, 2), c(4, 3)),
  seeds = c(30, 31),
  rlen = 40,
  k = 2:4
)
```

This design attempts 12 SOM fits: three subsamples, two grids and two random
starts. The small budget keeps the example quick. It is not a default for a
research analysis.

## Run the workflow

```{r workflow}
workflow <- run_som_workflow(
  dat,
  spec,
  resamples,
  preprocess = som_preprocess(
    "identity",
    center = TRUE,
    scale = TRUE
  ),
  cross_models = c("kmeans", "ward")
)
workflow
summary(workflow)
```

Preprocessing is learned within each subsample. Assessment rows therefore do
not influence the means or standard deviations used for training.

## Read the results in order

Start with fit success and map diagnostics.

```{r representation}
workflow$audit$grid_summary
plot(workflow$audit)
```

Quantization error, topographic error and empty-unit rate describe the map.
None of them tells us whether a hard partition is reproducible.

Next, inspect the candidate values of `k`.

```{r partitions}
workflow$partitions$stability
plot(workflow$partitions)
```

ARI and AMI measure agreement among repeated partitions. They are not
classification accuracy. Read them together with the number of successful
fits and the joint sample coverage.

Now look at sample-level support for one prespecified candidate. We use `k = 3`
because this teaching simulation has three generated classes.

```{r consensus}
consensus_k3 <- workflow$consensus$k3
consensus_k3$cluster_summary
plot(consensus_k3)
```

Low support or high assignment entropy points to uncertain membership. A
consensus label is a summary of repeated assignments, not a probability that
the label is correct.

Finally, compare the SOM partition with K-means and Ward.D2 on the same
eligible analysis splits.

```{r cross-model}
workflow$cross_comparison$summary
plot(workflow$cross_comparison)
```

Cross-model agreement is useful triangulation. It should not be turned into a
model ranking.

## Use known labels only after fitting

Because this is a simulation, we can compare the `k = 3` consensus with the
generated labels. Those labels were not used for training.

```{r external-labels}
evaluate_external_labels(consensus_k3)
```

In a field study, a recorded category may itself be uncertain or answer a
different biological question. Treat it as external evidence, not automatic
ground truth.

## What to save

Keep the following with any reported analysis:

- the input data and their provenance;
- stable sample identifiers and the sampling design;
- transformations, grids, seeds, iterations and candidate `k` values;
- warnings, failed fits and all evidence tables; and
- the package version and R session information.

```{r reproducibility}
packageVersion("SOMevidence")
sessionInfo()
```

Next, use the guided Shiny tutorial for an interactive introduction or the
resampling tutorial to match the workflow to a real monitoring design.
