---
title: "Using the guided Shiny app"
author: "Shaowen Ye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the guided Shiny app}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Start the app

The Shiny app is a practical way to learn the workflow and prepare an analysis.
It uses the same package functions as the R interface.

```{r launch, eval = FALSE}
library(SOMevidence)
shiny::runApp(launch_som_app())
```

The app helps you choose settings, checks the design before fitting and
explains each result view. For a publication analysis, export the R script and
continue in R. The R interface is better suited to large ensembles, multiple
data layers and custom sensitivity analyses.

Use **Language / 语言** to switch between English and Chinese. The current
data, parameters and computed results stay in place. The exported R script
remains in English so that it is portable across R sessions and publications.

Move the pointer over a dotted-underlined control label to see a one-sentence
explanation. Important result-table terms, including ARI and AMI, also have
short hover notes. Drag the vertical divider to change the width of the left
and right panels. Arrow keys make small adjustments when the divider has
keyboard focus, and a double-click restores the default width.

`SOMevidence` sends no telemetry. Files selected in a locally running app stay
on your computer. A remotely deployed app sends selected files to its server.

## Learn with the four examples

Choose a **Guided example** in the left panel. The app will load the data and
apply a short set of recommended settings.

| Example | Question | Resampling |
|---|---|---|
| Discrete classes | Can a clear partition be recovered repeatedly? | Rows |
| Continuous gradient | Is the map useful even when hard classes are weak? | Rows |
| Grouped monitoring | What happens when records are nested within sites? | Whole groups |
| Domain transfer | Does the map cover a monitoring domain left out of training? | Leave one domain out |

Use **Reset recommended settings** to undo changes. Use **Download example
CSV** if you want to inspect the table outside the app.

The columns `indicator_01` to `indicator_06` are generic simulated
measurements. They do not stand for fixed environmental variables. When you
upload a CSV, the app keeps your original variable names.

These are teaching examples. They are not performance benchmarks and their
settings are not general recommendations for field studies.

### What each example teaches

**Discrete classes** contains three generated groups. The group label is kept
out of training. This lets you see how map quality, partition stability,
sample support and cross-model agreement fit together.

**Continuous gradient** has no generated class boundary. It shows that a SOM
can represent continuous structure well without supporting a strong hard
classification.

**Grouped monitoring** contains repeated records from simulated sites. The app
resamples whole sites. This avoids treating related observations as new
independent information.

**Domain transfer** leaves out one simulated monitoring domain at a time. Its
outputs describe mapping distance and coverage, not prediction accuracy or the
cause of the domain difference.

## Upload your own CSV

Choose **Upload CSV**, then use **Download CSV template** as a starting point.
The names can change, but keep the roles clear.

| Role | Template column | Used to train the SOM? |
|---|---|---:|
| Sample identifier | `sample_id` | No |
| Sampling group | `sampling_group` | No |
| Sampling time | `survey_date` | No |
| Transfer domain | `reporting_region` | No |
| Measured variables | columns beginning with `measure_` | Yes |
| External label | `external_label` | No |

The app does not select uploaded predictors for you. Check every numeric
column. IDs, coordinates, dates, weights and known labels should not enter the
training matrix simply because they are numeric.

The selected predictors must be complete. The app does not hide imputation or
choose a missing-distance rule. If missing values are part of the study, plan
that analysis directly in R.

## Work down the left panel

### 1. Data and study design

Select the measured variables. Then identify the columns that describe sample
identity, grouping, time and transfer domains. These columns stay outside the
SOM feature space.

### 2. Preprocessing

Choose a transformation from the way the variables are measured and
distributed. Do not try several transformations and retain only the one that
gives the most attractive classes. Centering and scaling are estimated within
each analysis split.

### 3. Resampling design

Match the resampling unit to data collection:

- use `subsample` for independent rows;
- use `group_subsample` for sites, individuals or transects;
- use `leave_domain_out` for campaigns, regions or instruments; and
- use `full` only when you want repeated starts without data perturbation.

### 4. SOM ensemble

State the grid, random seeds, training iterations and candidate `k` values.
The app checks whether each split contains enough observations for the planned
map and partitions.

### 5. Triangulation and export

K-means and Ward.D2 are the default reference methods. GMM is optional. They
are fitted on the same eligible splits. Their role is to check whether similar
partition geometry appears under another algorithm, not to create a ranking.

Press **Run workflow** only after the preflight panel is ready.

## Read the results

| View | Ask | Do not conclude |
|---|---|---|
| Representation | Does the map represent the observations? | That discrete classes exist |
| Partition stability | Does the partition recur? | That stable groups are ecological truth |
| Consensus | Which sample assignments are well supported? | That a label is certainly correct |
| Cross-model agreement | Does similar geometry appear elsewhere? | Accuracy or a winning model |

The help box above each plot gives three short prompts: the question, what to
inspect and the interpretive limit.

Check the **Run diagnostics** table before interpreting a figure. Report failed
fits and warnings. For partition results, read ARI or AMI with joint sample
coverage. For consensus, inspect both membership support and assignment
entropy.

## If preflight stops the run

- **Select predictors:** choose measured variables and exclude design columns.
- **Predictors must be complete:** plan a justified missing-data analysis in R.
- **No planned SOM fit is eligible:** reduce the grid or revise the split.
- **Repeated analysis sets:** reduce repeats or change the resampling design.
- **No cross-model split is eligible:** review sample size and candidate `k`.

These messages prevent a run that cannot answer the stated question. They are
not software failures.

## Export a reproducible record

After a successful run, download the R script and the YAML snapshot. The R
script is executable. The YAML file is a readable record of the controls, but
the current app cannot import it.

For an uploaded dataset, place the CSV in the `data` directory expected by the
script. Run the script in a clean R session. Save the input provenance, package
version, warnings, failures and results with the final analysis.

Move to the R interface when you need multiple layers, missing-data
sensitivity, time blocks, a larger computing budget, study-specific decision
rules or automated reporting.
