---
title: "Choose the right resampling unit"
author: "Shaowen Ye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choose the right resampling unit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why rows are not always independent

Monitoring tables often contain several records from the same site,
individual, transect or campaign. If we split those rows independently, closely
related records can appear in both the analysis and assessment sets. The
partition may then look more reproducible than it is for a new sampling unit.

The solution is simple: resample the unit that was independently sampled.

## Compare row and site resampling

The example below creates repeated observations within 20 simulated sites.
Generated class labels are stored as metadata and are not used for training.

```{r grouped-data}
library(SOMevidence)

grouped <- simulate_som_scenario(
  "grouped_pseudoreplication",
  n = 120,
  p = 6,
  n_groups = 20,
  group_icc = 0.7,
  seed = 101
)
grouped
```

We now create two designs. The first samples rows. The second samples whole
sites.

```{r resampling-designs}
row_splits <- som_resamples(
  grouped,
  method = "subsample",
  repeats = 3,
  prop = 0.8,
  seed = 102
)

site_splits <- som_resamples(
  grouped,
  method = "group_subsample",
  repeats = 3,
  prop = 0.8,
  seed = 102
)

spec <- som_spec(
  grids = list(c(3, 2), c(4, 3)),
  seeds = c(103, 104),
  rlen = 40,
  k = 2:4
)
```

Only the resampling unit changes.

```{r grouped-workflows}
row_result <- run_som_workflow(
  grouped,
  spec,
  row_splits,
  cross_models = c("kmeans", "ward")
)

site_result <- run_som_workflow(
  grouped,
  spec,
  site_splits,
  cross_models = c("kmeans", "ward")
)

row_result$partitions$stability
site_result$partitions$stability
```

Treat this as a design sensitivity, not a competition between algorithms. If
the results differ, the sampling hierarchy matters. The field design tells you
which result answers the scientific question.

## Leave out a monitoring domain

Sometimes the question is not about new sites within the same survey. We may
want to know how a map behaves in a different season, region, campaign or
instrument domain.

```{r transfer-data}
shifted <- simulate_som_scenario(
  "gradient",
  n = 120,
  p = 6,
  n_domains = 3,
  domain_shift = 1,
  seed = 110
)

domain_splits <- som_resamples(
  shifted,
  method = "leave_domain_out",
  domain = "domain"
)

domain_ensemble <- fit_som_ensemble(
  shifted,
  som_spec(
    c(4, 3),
    seeds = c(111, 112),
    rlen = 40,
    k = 2:4
  ),
  domain_splits
)

transfer <- audit_transfer(domain_ensemble)
transfer$metrics
```

```{r transfer-plot, fig.width = 6, fig.height = 4}
plot(transfer)
```

The distance and occupancy measures show whether the held-out domain maps like
the training data. They are not accuracy measures. This simulation mixes range
extrapolation with an added domain shift, so it also cannot identify a causal
mechanism.

## Map a later batch without refitting

An operational workflow may freeze an ensemble and map a later batch. The new
data need the same named layers and variables. Each map reuses the
preprocessing learned from its own training data.

```{r new-batch}
training_rows <- shifted$metadata$domain != "domain_03"

training <- som_data(
  layers = lapply(
    shifted$layers,
    function(x) x[training_rows, , drop = FALSE]
  ),
  id = shifted$metadata$id[training_rows]
)

new_batch <- som_data(
  layers = lapply(
    shifted$layers,
    function(x) x[!training_rows, , drop = FALSE]
  ),
  id = shifted$metadata$id[!training_rows],
  domain = shifted$metadata$domain[!training_rows]
)

frozen_ensemble <- fit_som_ensemble(
  training,
  som_spec(
    c(4, 3),
    seeds = c(120, 121),
    rlen = 40,
    k = 2:4
  ),
  keep_models = TRUE
)

new_mapping <- map_som_ensemble(frozen_ensemble, new_batch)
new_mapping$summary
```

```{r new-batch-plot, fig.width = 6, fig.height = 4}
plot(new_mapping)
```

Best-matching unit numbers belong to a specific fitted map. Do not compare a
node number across different grids as if it were the same class.

## Define time or space blocks yourself

Use `block_subsample` when the study design already defines seasons, campaigns,
reaches or time windows. Pass those blocks through `unit`. The package does not
guess a block length from row order.

The practical rule is: identify the independent sampling unit first, then
choose the resampling method.
