---
title: "Read the maps and uncertainty plots"
author: "Shaowen Ye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Read the maps and uncertainty plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Fit a small example

A component plane shows one fitted map. It cannot show how much the result
changes across resampled data, grid sizes and random starts. We therefore read
the map together with the ensemble plots.

```{r fit}
library(SOMevidence)

dat <- simulate_som_scenario(
  "clusters",
  n = 120,
  p = 6,
  seed = 201
)

resamples <- som_resamples(
  dat,
  method = "subsample",
  repeats = 3,
  seed = 202
)

spec <- som_spec(
  grids = list(c(4, 3), c(5, 4)),
  seeds = c(203, 204),
  rlen = 50,
  k = 2:5
)

workflow <- run_som_workflow(
  dat,
  spec,
  resamples,
  cross_models = c("kmeans", "ward")
)

fit_id <- workflow$ensemble$fits[[1]]$id
```

## Look at one map

```{r component-plane, fig.width = 8, fig.height = 4.5}
plot_som_plane(
  workflow$ensemble,
  fit_id = fit_id,
  variables = colnames(dat$layers$environment)[1:4]
)
```

The colours show codebook values after the preprocessing used for this fit.
The subtitle identifies the ensemble member. If more than one map is available,
`fit_id` is required so that the displayed map is unambiguous.

Two companion views help explain the map:

```{r map-diagnostics, fig.width = 8, fig.height = 3.8}
plot_som_plane(
  workflow$ensemble,
  type = "occupancy",
  fit_id = fit_id
)

plot_som_plane(
  workflow$ensemble,
  type = "neighbour_distance",
  fit_id = fit_id
)
```

Occupancy shows how many observations use each map unit. Neighbour distance
shows changes between adjacent codebook vectors. Neither view tells us whether
a hard classification is stable.

## Put ensemble evidence beside the map

```{r audit-plots, fig.width = 7, fig.height = 4}
plot(workflow$audit)
plot(workflow$partitions)
```

The first plot summarizes representation diagnostics and fit success. The
second shows how candidate partitions vary across the ensemble. Read the
displayed intervals and coverage rather than choosing a result from colour or
visual separation alone.

## Find uncertain samples

```{r consensus, fig.width = 6, fig.height = 4}
consensus_k3 <- workflow$consensus$k3
plot(consensus_k3)
consensus_k3$cluster_summary
```

Membership support and assignment entropy show where the repeated fits agree
and where they do not. A consensus label is still conditional on the chosen
data, preprocessing, ensemble and `k`.

## Compare reference methods

```{r cross-model, fig.width = 7, fig.height = 4}
plot(workflow$cross_comparison)
```

This plot shows agreement with the prespecified reference methods. It does not
rank the methods or turn agreement into accuracy.

For a report, keep at least three views together: one clearly identified map,
the ensemble representation diagnostics and the partition or consensus
evidence. This is more informative than a single attractive SOM figure.
