---
title: "Worked example: Palmer penguin morphology"
author: "Shaowen Ye"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked example: Palmer penguin morphology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_tutorial_dependencies <-
  requireNamespace("palmerpenguins", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_tutorial_dependencies
)
```

## The question

This example asks how reproducibly a SOM represents and partitions four penguin
body measurements. Species is kept as an external label and is inspected only
after fitting.

The example does not assume that a SOM class is a species. Morphology overlaps
among species and does not define taxonomy by itself.

The `palmerpenguins` package distributes the data under CC0. Its documentation
provides the original data source and citations. If the optional package is not
installed, this vignette displays the code without running it.

## Prepare the data

```{r prepare-data}
library(SOMevidence)

raw <- palmerpenguins::penguins
variables <- c(
  "bill_length_mm",
  "bill_depth_mm",
  "flipper_length_mm",
  "body_mass_g"
)

keep <- stats::complete.cases(
  raw[, c(variables, "species", "year")]
)
penguins <- raw[keep, , drop = FALSE]
sample_ids <- sprintf("penguin_%03d", seq_len(nrow(penguins)))

morphology <- som_data(
  x = penguins[, variables],
  id = sample_ids,
  time = penguins$year,
  domain = factor(penguins$year),
  external_label = stats::setNames(
    as.character(penguins$species),
    sample_ids
  )
)
morphology
```

Stable IDs link results back to observations. Year defines the transfer domain.
Species remains outside the training matrix.

## Leave out one year at a time

Each split trains on two sampling years and assesses the third. Preprocessing
is estimated from the training years only.

```{r design}
year_splits <- som_resamples(
  morphology,
  method = "leave_domain_out",
  domain = "domain"
)

spec <- som_spec(
  grids = list(c(4, 3), c(5, 4)),
  seeds = c(41, 42),
  rlen = 50,
  k = 2:4
)

year_splits
```

The ensemble is intentionally small so that the vignette builds quickly.

## Fit and inspect the workflow

```{r workflow}
workflow <- run_som_workflow(
  morphology,
  spec,
  year_splits,
  preprocess = som_preprocess(
    center = TRUE,
    scale = TRUE
  ),
  cross_models = c("kmeans", "ward")
)
summary(workflow)
```

Read the evidence streams separately.

```{r summaries}
workflow$audit$grid_summary
workflow$partitions$stability
workflow$consensus$k3$cluster_summary
workflow$cross_comparison$summary

transfer <- audit_transfer(workflow$ensemble)
transfer$metrics
```

```{r evidence-plots, fig.width = 7, fig.height = 4, out.width = "90%"}
plot(workflow$audit)
plot(workflow$partitions)
plot(transfer)
```

The map audit concerns representation. The partition plot concerns repeated
hard classifications. The transfer plot concerns the held-out year. None is a
classification-accuracy plot.

## Compare with species after fitting

For this teaching example, inspect `k = 3` because the data contain three
recorded species. This is a choice made before viewing the agreement results.
It is not evidence that `k = 3` is the best unsupervised partition.

```{r external-labels}
consensus_k3 <- workflow$consensus$k3
external <- evaluate_external_labels(consensus_k3)

external
external$contingency
external$composition
```

ARI, AMI and the contingency table describe agreement among the evaluable
penguins. They do not prove that the partition is biologically correct. The
main lesson is procedural: keep labels out of unsupervised training, then bring
them back transparently for interpretation.

## Save the record

Keep the exact data preparation, IDs, transfer design, SOM settings, warnings
and failed fits with the result.

```{r reproducibility}
packageVersion("SOMevidence")
packageVersion("palmerpenguins")
sessionInfo()
```
