---
title: "Reproducibility and API versions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducibility and API versions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why versions matter

An analysis can change when a function argument, default or returned field
changes. The function name may stay the same. `SOMevidence` therefore keeps
machine-readable records of its public functions and result objects.

These records serve two purposes:

- package tests can detect an unintended interface change; and
- analysts can see which structure belongs to a released version.

They do not guarantee identical numerical results when an analysis relies on
defaults that later change. For a published workflow, record all important
settings directly.

## Inspect the current contracts

```{r contracts}
contract_path <- function(name) {
  candidates <- c(
    file.path("inst", "extdata", name),
    file.path("..", "inst", "extdata", name),
    system.file("extdata", name, package = "SOMevidence")
  )
  available <- candidates[nzchar(candidates) & file.exists(candidates)]
  if (!length(available)) {
    stop("The requested API contract is not available: ", name)
  }
  available[[1L]]
}

api_contract <- utils::read.csv(
  contract_path("api-contract-v1.2.0.csv"),
  check.names = FALSE
)

api_formals <- utils::read.csv(
  contract_path("api-formals-v1.2.0.csv"),
  check.names = FALSE
)

object_contract <- utils::read.csv(
  contract_path("object-contract-v1.2.0.csv"),
  check.names = FALSE
)

table(api_contract$lifecycle)
head(api_formals)
head(object_contract[c("class", "lifecycle")])
```

The API contract lists exported functions and their lifecycle. The formals
contract records arguments and defaults. The object contract records the
fields that downstream code may expect from public results.

The version 1.2.0 object contract also gives each result a
`som_contract_version`. It can align external labels by sample ID and keeps
missing, excluded and insufficiently replicated records separate.

Release history and changes to defaults belong in `NEWS.md`. This tutorial
focuses on how to use the contracts rather than repeating the changelog.

## Stable and experimental functions

Most functions used for data preparation, resampling, SOM fitting, audits and
transfer are stable. Their established arguments and documented result fields
are maintained within a major version.

Some interfaces remain experimental:

```{r experimental}
api_contract[
  api_contract$lifecycle == "experimental",
  c("function", "return_class", "notes")
]
```

Experimental does not mean untested. It means that feedback from wider use may
still change the interface or the way results are organized. In version 1.2.0,
this applies to the Shiny app, simulation helpers, sensitivity workflow and
continuous cross-fit representation audit.

The Shiny app can export a runnable R script. Its YAML file is a configuration
snapshot, not an importable analysis file. Keep the script with the data,
software versions, warnings and results.

## Each result answers one question

The package uses separate object classes so that different types of evidence
are not confused:

| Object | Question |
|---|---|
| `som_audit` | How well did the maps represent the analysis data? |
| `som_representation_audit` | How similar was sample topology across selected maps? |
| `som_partitions` | How reproducible were candidate hard partitions? |
| `som_consensus` | Which sample assignments were repeatedly supported? |
| `som_cross_comparison` | Did reference methods produce similar partitions? |
| `som_transfer_audit` | How did held-out domains map to trained SOMs? |
| `som_external_assessment` | How did a consensus agree with labels excluded from training? |
| `som_defensibility` | Did an analyst's stated evidence requirements pass? |

No object is a certificate of ecological truth. A `supported` decision means
that the stated checks passed for that workflow. It does not prove that nature
contains the displayed classes.

## Open an older saved object

`upgrade_som_object()` can add fields when the migration is deterministic and
the older object already contains the required evidence.

```{r upgrade, eval = FALSE}
old_result <- readRDS("saved-som-result.rds")
current_result <- upgrade_som_object(old_result)
```

If the older result lacks required evidence, refit the analysis. Do not invent
missing fields or copy them from another run.

## Record a publication analysis

Save:

- the package and R versions;
- input provenance and stable sample IDs;
- preprocessing and resampling choices;
- grids, seeds, iterations and candidate `k` values;
- reference methods and any decision rules; and
- complete warnings, failures and evidence tables.

```{r report-version}
packageVersion("SOMevidence")
sessionInfo()
```

The contract tells software what a result contains. The analysis record tells
readers how that result was produced. A reproducible study needs both.
